Approximate Bayesian Computation in Large Scale Structure: constraining the galaxy-halo connection

### 2017-03-08 ccppabc/text/revision/main.tex
directory clean up; significant edits to Section 3.4 to generally address Ref Report #1; see green text for issues requiring attention

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols

@@

@@

\definecolor{dred}{rgb}{0.75,0,0}
\definecolor{darkred}{rgb}{0.5,0,0}
\definecolor{darkgreen}{rgb}{0,0.5,0}

@@

%%%%%%%%%%%%%%%%%%%%%%% SECTION: %%%%%%%%%%%%%%%%%%%%%%%
% SECTION: Comparison to MCMC analysis  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparison to the Gaussian Likelihood MCMC Analysis} \label{sec:abcvsmcmc}
In order to assess the quality of the parameter inference described in the previous section, 
we compare the ABC-PMC results with the HOD parameter constraints assuming a Gaussian likelihood 
function. The model used for the Gaussian likelihood analysis is different than the forward 
generative model adopted for the ABC-PMC algorithm, to be consistent with the standard approach.
In the ABC analysis, the model accounts for sample variance by randomly sampling a subvolume to be 
populated with galaxies. The Gaussian likelihood analysis assumes a covariance matrix that captures 
the uncertainties from the sample variance, and therefore in the model, we populate the halos of
the entire $\mathtt{MultiDark}$ simulation rather than a subvolume.

We introduce the observable $\mathbf{x}$ as a combination of the summary statistics of the 
galaxy mock catalog used in the inference. When we use $\ngalaxy$ and $\xigg(r)$ as observables, 
$\mathbf{x} = [\ngalaxy , \xigg]$, while $\mathbf{x} = [\ngalaxy , \gmf]$ when we use $\ngalaxy$ and $\gmf(N)$
Based on this notation, we write likelihood function as 
\begin{eqnarray}
-2 \ln \mathcal{L}(\theta | d) &=& \Delta \mathbf{x}^{T}\widehat{C^{-1}}\Delta \mathbf{x} + \ln\Big[(2\pi)^{d}\mathrm{det}(C)\Big], \label{loglike}\\
\Delta \mathbf{x} &=& [\mathbf{x}_{obs} -\mathbf{x}_{mod}],
\end{eqnarray}
where $\Delta \mathbf{x}$ is the difference between the measured $\mathbf{x}_{obs}$ of the mock observations 
and that of the model $\mathbf{x}_{mod}(\mathbb{\theta})$. $d$ is the dimension of the observable $\mathbf{x}$. 
When we use $\mathbf{x} = [\ngalaxy, \xigg]$, $d = 13$. When we use $\mathbf{x} = [\ngalaxy, \gmf]$, $d = 10$.  
$\widehat{C^{-1}}$ is the estimate of the inverse covariance matrix, which we calculate 
following  \cite{hartlap2007}:
\beq
\widehat{C^{-1}} = \frac{N_{\rm mocks}-d - 1}{N_{\rm mocks} -1}\widehat{C}^{-1},
\eeq
where $\widehat{C}$ is the estimated covariance matrix and $N_{\rm mock}$ is the number of mocks used to 
estimate $\widehat{C}$. $N_{\rm mock} = 124$ (Section \ref{sec:mock_obv}). Depending on $\mathbf{x}$, 
$\widehat{C}$ is given by the appropriate block of the covariance matrix Eq. \ref{eq:cov} corresponding 
to the observables. We note that in the covariance matrix the dependence on the HOD parameters is neglected. 
Therefore, the second term in the expression of the log-likelihood (Eq. \ref{loglike}) can be neglected. 
We sample from the posterior distribution of the likelihood given the prior distribution using the 
MCMC sampler $\mathtt{EMCEE}$ (\citealt{emcee}).

In Figure \ref{fig:hist_nbarxi} we compare the ABC-PMC and Gaussian likelihood MCMC 
results for the marginalized posterior PDFs over three parameters of the HOD model $\{\log \mathcal{M}_{\rm min},
\alpha , \log\mathcal{M}_1\}$ using $\ngalaxy$ and $\xigg$ as observables. 
Figure \ref{fig:hist_nbargmf} makes the same comparison when $\ngalaxy$ and $\gmf$ are used in the inference.

Figure \ref{fig:hist_nbarxi} demonstrates in the top panel that constraints obtained from the two 
methods are consistent with the ``true'' HOD parameters represented by the vertical dashed lines. 
The marginalized posteriors over the parameters follow each other very closely. The lower panel of 
Figure \ref{fig:hist_nbarxi} plot the $68\%$ and $95\%$ confidence intervals of the constraints 
derived from the two methods as a box plot. The uncertainty over the parameter $\log \mathcal{M}_\mathrm{min}$ 
is slightly smaller while the uncertainties of the parameters $\alpha$ and $\log M_{1}$ 
are slightly larger for our ABC-PMC results.

In Figure \ref{fig:hist_nbargmf}, when we use $\ngalaxy$ and $\gmf$ as observables, both methods 
produce consistent constraints with each other and ``true'' HOD parameters. The lower panel of Figure 
\ref{fig:hist_nbargmf} shows that the uncertainties from the two methods are comparable. The 
ABC-PMC constraints over $\alpha$ is slightly less biased and slightly less precise.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABC vs MCMC Histograms 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
%\begin{center}
\includegraphics[width=1.\textwidth]{figs/paper_ABCvsMCMC_nbarxi.pdf}
\caption{\label{fig:hist_nbarxi} We compare the $\log \mathcal{M}_{\rm min}$, $\alpha$, and $\log \mathcal{M}_{1}$ parameter constraints from ABC-PMC (orange) to the constraints from the Gaussian Likelihood MCMC (blue) using $\ngalaxy$ and $\xigg(r)$ as observables. The \emph{top} panels show the histograms of the marginalized posterior PDFs over the parameters. In the \emph{bottom} panels, we include box plots marking the confidence intervals of the posterior distributions. The boxes represent the $68\%$ confidence interval while the ``whiskers'' represent the $95\%$ confidence interval. We also
plot the ``true'' HOD parameters with vertical black dashed line. Marginalized posterior PDFs 
obtained from the two methods are consistent with each other. The ABC-PMC constraints are slightly 
narrower for $\log \mathcal{M}_{\rm min}$, slightly wider for $\log \mathcal{M}_{1}$, and slightly 
less biased for $\alpha$.} 
%\end{center}
\end{figure*}

>>>>

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{xcolor}	% Extra maths symbols

@@

\definecolor{dred}{rgb}{0.75,0.,0.}
\definecolor{darkred}{rgb}{0.5,0.,0.}
\definecolor{darkgreen}{rgb}{0.,0.5,0.}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%% SECTION: %%%%%%%%%%%%%%%%%%%%%%%
% SECTION: Comparison to MCMC analysis  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparison to the Gaussian {\color{dred} Pseudo - Likelihood} MCMC Analysis} \label{sec:abcvsmcmc}
In order to assess the quality of the parameter inference described in the previous section, 
we compare the ABC-PMC results with the HOD parameter constraints 
{\bf \color{dred} from}
assuming a Gaussian likelihood 
function. The model used for the Gaussian likelihood analysis is different than the forward 
generative model adopted for the ABC-PMC algorithm, to be consistent with the standard approach.

In the ABC analysis, the model accounts for sample variance by randomly sampling a subvolume to be 
populated with galaxies. 
{\bf \color{dred} %%%
Instead, in the Gaussian pseudo-likelihood analysis, the covariance matrix is assumed to capture the
uncertainties from sample variance. Hence, in the model for the Gaussian pseudo-likelihood analysis, 
we populate halos of the {\em entire} $\mathtt{MultiDark}$ simulation rather than a subvolume.
We describe the Gaussian pseudo-likelihood analysis below.
}
%The Gaussian likelihood analysis assumes a covariance matrix that captures the uncertainties from the sample variance, and therefore in the model, we populate the halos of the entire $\mathtt{MultiDark}$ simulation rather than a subvolume.

{\bf \color{dred}
To write down the Gaussian pseudo-likelihood, we first introduce the vector $\mathbf{x}$: 
a combination of the summary statistics (observables) for a galaxy catalog. 
When we use $\ngalaxy$ and $\xigg(r)$ as observables in the analysis: $\mathbf{x} = [\ngalaxy , \xigg]$;
when we use $\ngalaxy$ and $\gmf(N)$ as observables in the analysis: $\mathbf{x} = [\ngalaxy , \gmf]$.
Based on this notation, we can write pseudo-likelihood function as 
\begin{eqnarray}
-2 \ln \mathcal{L}(\theta | d) &=& \Delta \mathbf{x}^{T}\widehat{C^{-1}}\Delta \mathbf{x} + \ln\Big[(2\pi)^{d}\mathrm{det}(C)\Big], \label{loglike}
\end{eqnarray}
where 
\begin{eqnarray}
\Delta \mathbf{x} &=& [\mathbf{x}_{obs} -\mathbf{x}_{mod}], 
\end{eqnarray}
the difference between $\mathbf{x}_{obs}$, measured from the mock observation, 
and $\mathbf{x}_{mod}(\mathbb{\theta})$ measured from the mock catalog generated 
from the model with parameters $\theta$ .
$d$ here is the dimension of $\mathbf{x}$ (for $\mathbf{x} = [\ngalaxy, \xigg]$, $d = 13$; 
for $\mathbf{x} = [\ngalaxy, \gmf]$, $d = 10$).  
$\widehat{C^{-1}}$ is the inverse covariance matrix, which we estimate following \cite{hartlap2007}:
\begin{eqnarray}
\widehat{C^{-1}} = \frac{N_{\rm mocks}-d - 1}{N_{\rm mocks} -1} \; \widehat{C}^{-1}.
\end{eqnarray}
$\widehat{C}$ is the estimated covariance matrix, calculated using the corresponding 
$\mathbf{x}$ block of the covariance matrix from Eq.~\ref{eq:cov}, and $N_{\rm mock}$ 
is the number of mocks used for the estimation ($N_{\rm mock} = 124$; see Section \ref{sec:mock_obv}).
We note that in $\widehat{C}$ the dependence on the HOD parameters is neglected, 
so the second term in the expression of Eq.~\ref{loglike} can be neglected. 
Finally, using this pseudo-likelihood, we sample from the posterior distribution 
given the prior distribution using the MCMC sampler $\mathtt{emcee}$ (\citealt{emcee}). 
}

{\bf \color{dred}
In Figures~\ref{fig:hist_nbarxi} and \ref{fig:hist_nbargmf}, we compare the results from
ABC-PMC and Gaussian pseudo-likelihood MCMC analyses using $[\ngalaxy, \xigg]$ and 
$[\ngalaxy, \gmf]$ as observables respectively. The top panels in each figure 
compares the marginalized posterior PDFs for three parameters of the HOD model: 
$\{\log \mathcal{M}_{\rm min}, \alpha , \log\mathcal{M}_1\}$. The lower panels in each 
figure compares the $68\%$ and $95\%$ confidence intervals of the constraints 
derived from the two inference methods as a box plot. The ``true'' HOD parameters
are marked by vertical dashed lines in each panel. 
}

{\bf \color{dred}
In Figures~\ref{fig:hist_nbarxi} and \ref{fig:hist_nbargmf}, for each of the parameters, 
the marginalized posteriors from both inference methods are comparable. Similarly, 
the lower panels of the figures reveal that the constraints derived for each parameter
from the two methods are also consistent, with minor discrepancies. Furthermore, we 
find that both methods produce constraints consistent with the ``true'' HOD parameters.
}
{\bf \color{darkgreen}
CHH: A few sentences discussing the width of the distributions for each of the panels
have been commented out. Not sure if it's meaningful to include such slight discrepancies. 
@MJV: Could you take a look at what's written above and see if you feel that it's sufficient?  
}
%The uncertainty over the parameter $\log \mathcal{M}_\mathrm{min}$ is slightly smaller while the uncertainties of the parameters $\alpha$ and $\log M_{1}$ are slightly larger for our ABC-PMC results.  In Figure \ref{fig:hist_nbargmf}, when we use $\ngalaxy$ and $\gmf$ as observables, both methods produce consistent constraints with each other and ``true'' HOD parameters. The lower panel of Figure \ref{fig:hist_nbargmf} shows that the uncertainties from the two methods are comparable. The ABC-PMC constraints over $\alpha$ is slightly less biased and slightly less precise.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABC vs MCMC Histograms 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
%\begin{center}
\includegraphics[width=1.\textwidth]{figs/paper_ABCvsMCMC_nbarxi.pdf}
\caption{\label{fig:hist_nbarxi} 
{\bf \color{dred}
We compare the $\log \mathcal{M}_{\rm min}$, $\alpha$, and $\log \mathcal{M}_{1}$ parameter 
constraints from ABC-PMC (orange) to constraints from the Gaussian pseudo-ikelihood MCMC (blue) 
using $\ngalaxy$ and $\xigg(r)$ as observables. The \emph{top} panels compares the two methods' 
marginalized posterior PDFs over the parameters. In the \emph{bottom} panels, we include box 
plots marking the confidence intervals of the posterior distributions. The boxes represent the 
$68\%$ confidence interval while the ``whiskers'' represent the $95\%$ confidence interval. We mark 
the ``true'' HOD parameters with vertical black dashed line. The marginalized posterior PDFs 
}
obtained from the two methods are consistent with each other. The ABC-PMC constraints are slightly 
narrower for $\log \mathcal{M}_{\rm min}$, slightly wider for $\log \mathcal{M}_{1}$, and slightly 
less biased for $\alpha$.} 
%\end{center}
\end{figure*}
